Model Finding for (Co)recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples Utilisables par Isabelle et Coq'' (CUIC). }

### 2015-05-05 tex/main.tex
tuned typesetting

%%%%%%%%%%%%%%%%%%%%%%% file typeinst.tex %%%%%%%%%%%%%%%%%%%%%%%%%
%
% This is the LaTeX source for the instructions to authors using
% the LaTeX document class 'llncs.cls' for contributions to
% the Lecture Notes in Computer Sciences series.
% http://www.springer.com/lncs       Springer Heidelberg 2006/05/04
%
% It may be used as a template for your own input - copy it
% to a new file with a new name and use it as the basis
% for your article.
%
% NB: the document class 'llncs' has its own and detailed documentation, see
% ftp://ftp.springer.de/data/pubftp/pub/tex/latex/llncs/latex2e/llncsdoc.pdf
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[runningheads,a4paper]{llncs}
\usepackage{mathptmx}
\usepackage{listings}
\usepackage{colonequals}
%\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{framed}

@@

@@

@@

\newcommand{\boolop}{\oplus}
\newcommand{\forallf}[1]{\forall_{#1}}
\newcommand{\fnull}{\emptyset}
\newcommand{\farg}[1]{a_{#1}}
\newcommand{\vecfarg}[1]{\vec{a}_{#1}}
\newcommand{\fargsort}[1]{\upsilon_{#1}}

@@

\newcommand{\euf}{\ensuremath{\mathrm{E}}\xspace}
\newcommand{\ari}{\ensuremath{\mathrm{A}}\xspace}

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate $\teq$ of type $\tau \times \tau \to \Bool$
for each $\tau \in \ssorts{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite$ of type $\Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$ and
$\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each $\tau \in \ssorts{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in $\I$,
and each $\con{f}$ of type $\tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$ 
to a total function $f^\I$ of type $\tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where 
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is 
\emph{$T$-satisfiable} (resp., \emph{$T$-unsatisfiable}) 
if it is satisfied by some (resp., no) interpretation in $\Mo$.
Formulas $\varphi$ and $\psi$ are \emph{equisatisfiable} if
$\varphi$ is $T$-satisfiable if and only if $\psi$ is $T$-satisfiable.
For a given signature $\Sigma$
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the sort and function symbols in this theory as \emph{uninterpreted}.

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form $\forallf{f} \vec x. \varphi$, 
where $f$ is an uninterpreted function from $\sfuns{\Sigma}$, or the distinguished constant $\fnull$.
An annotated quantified formula $\forallf{f} \vec x. \varphi$ is a \emph{function definition (for $f$)} if 
$f$ is a non-$\fnull$ function of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
$\vec x$ is a tuple of $n$ variables of type $\tau_1, \ldots, \tau_n$
and $\forallf{f} \vec x. \varphi$ is equivalent to $\forall \vec x. f( \vec x ) \teq t$ for some $t$.
We will write $\forall \vec x. \varphi$ as shorthand for $\forallf{\fnull} \vec x. \varphi$,
and $\exists \vec x. \varphi$ as shorthand for $\neg \forall \vec x. \neg \varphi$.

\begin{definition}
A formula $\varphi$ is \emph{well-defined with respect to functions} $\{ f_1, \ldots, f_n \}$ if it is of the form:
\begin{eqnarray} \label{eq:wdf}
(\forallf{f_1} \vec x_1. \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n. \varphi_n) \wedge \psi
\end{eqnarray}
where $f_1, \ldots, f_n$ are not $\fnull$ and distinct, 
and $\psi$ contains no function definitions.
We will commonly call $\psi$ the \emph{conjecture} of $\varphi$.
\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
\end{definition}

@@

A set of function definitions $\{ \varphi_1, \ldots, \varphi_n \}$ 
is \emph{well-specified (in $T$)} if $\varphi_1 \wedge \ldots \wedge \varphi_n$ is $T$-satisfiable.
For instance, $\{ \forallf{f} \vec x. f( x ) \teq 0 \}$ is well-specified,
whereas $\{ \forallf{f} x. f( x ) \teq f( x ) + 1 \}$ is not. 
Note that a well-specified function definition need not specify a terminating function, e.g.
$\{ \forallf{f} x^\sortint. f( x ) \teq f( x - 1 ) \}$ or even $\{ \forallf{f} x. f( x ) \teq f( x ) \}$ are well-specified.
\ajr{revise}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instantiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For instance, the formula $\forall x\typ{\sortint}. f( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $f$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For instance, 
finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau}. \neg x \teq y \vee f( x ) \teq f( y )$
where the uninterpreted type $\tau$ is interpreted as a finite set, and $f$ as injective finite map.

Unfortunately, these fragments do not coincide with the vast majority of quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x\typ{\sortint}. \lite( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) )$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$, 
and the set of function definitions corresponding to $\sfundefs{\Sigma}$ are well-specified.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature $\extendsig{ \Sigma }$,
and a $\extendsig{ \Sigma }$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is to use an uninterpreted type \ct{to be defined} to abstract the set of \emph{relevant} tuples for each recursive function $f$,
and restrict the quantification of the function definition for $f$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $f$ are the ones for which the interpretation of $f( \vec t )$ is relevant to the satisfiability of $\varphi$.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{f} \vec x. t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $f \in \sfundefs{\Sigma}$
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{f}}. ( r_1 [ \vecfarg{f}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x. r_1, \forall \vec x. D_1 )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to a set of well-specified function definitions $\sfundefs{\Sigma}$.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an interpreted predicate.
}
\label{fig:encoding}
\end{figure}

In more detail, 
say our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $f \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted type $\fargsort{f}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ of type $(\fargsort{f} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{f} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{f}$ will denote the elements (tuples) on which the function $f$ is applied.
%The role of the uninterpreted functions $\farg{f}^1$, $\ldots$, $\farg{f}^n$ will be discussed more in the following.

Given the extended signature, we run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
This calls the subprocedure $\conv_0$, which takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas.
%The role of $D$ is to ensure that the (restricted) function definition for $f$ includes certain tuples in its domain, which we explain more in the following.
At a high level, the translation modifies $\varphi$ in two ways.
First, it restricts the quantification on function definitions for $f$ to the corresponding uninterpreted type $\fargsort{f}$.
Second, it augments $\varphi$ with additional constraints of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$
which ensure that this restriction for the defintion $f$ covers all relevant tuples of terms,
namely the ones for which an $f$-application exists and is relevant to the satisfiability of the current branch of $\varphi$.

In the case that $t$ is an application of (interpreted) predicate $\boolop$ (e.g. a boolean connective, or equality),
we recursively call $\conv$ on its subchildren $t_i$ and polarity $\pol( \boolop, i, p )$, where $\pol$ is defined as\footnote{The negation $\neg p$ of polarity $p$ is 
$\pfalse$ if $p$ is $\ptrue$,
$\ptrue$ if $p$ is $\pfalse$, and
$\pnone$ if $p$ is $\pnone$.}:
\begin{equation*}
\pol( \boolop, i, p ) = \begin{cases}
                         \pnone & \boolop \equiv \teq \text{ or } ( \boolop \equiv \lite \text{ and } i=0 ) \\
                         \neg p & \boolop \equiv \neg \\
                         p & \text{otherwise}
                         \end{cases}
\end{equation*}
If $t$ has positive polarity,
we construct a conjunction of the result of the recursive calls with formulas $D_1, \ldots, D_n$.
Similarly, if $t$ has negative polarity,
we construct a disjunction of the result of the recursive calls with the negation of formulas $D_1, \ldots, D_n$.
It $t$ has no polarity,
then we return the union of $D_1 \cup \ldots \cup D_n$.

In the case that $t$ is a quantified formula $\forallf{f} \vec x. t_1$, it recursively calls $\conv$ on its body with the same polarity.
In the case that $t$ is a function definition of a function in $\sfundefs{\Sigma}$,
we instead construct a quantified formula over a single variable $y$ of type $\fargsort{f}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{f}( y )$.
In this case, since function definitions are a top-level conjunct, by case analysis on the return values of $\conv$, we know that $D_1$ is empty.
In the case that $t$ is not a function definition, 
the resulting quantified formula is reconstructed with the term $r_1$ returned by the recursive call,
and a quantified prefix is appended to the formula $D_1$ returned by the recursive call.

Otherwise, $t$ is either an application of an uninterpreted predicate, or a term of a type other than $\Bool$.
In this case, the second component of our return value contains a constraint of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$
for each subterm $f( \vec t )$ of $t$ such that $f$ is a function in $\sfundefs{\Sigma}$.
When such constraints are asserted positively, this indicates that the tuple of uninterpreted functions $\vecfarg{f}$ must include $\vec t$ in its range.

@@

%\begin{lemma}
%Let $\M$ be a model satisfying $\conv( \varphi )$, and
%%and let $\conv( \varphi )^\ast$ be the quantifier-free formula obtained by 
%let $M$ be a set of literals such that $\M \models M$ and $M \models_p \conv( \varphi )$.
%For each assignment $\sigma$ over $\vars{ M }$,
%each $f( \vec t ) \in \terms( M \sigma )$ where $f \in \sfundefs{\Sigma}$, 
%there exists a formula of the form $\exists z\typ{\fargsort{f}}. ( \vecfarg{f}( z ) \teq \vec t )$ in $M \sigma$.
%\end{lemma}

>>>>

\documentclass[runningheads,a4paper]{llncs}
\usepackage[T1]{fontenc}
\usepackage[scaled=.82]{beramono}
\usepackage[scaled=.86]{helvet}
\usepackage{mathptmx}
\usepackage{listings}
\usepackage{colonequals}
%\usepackage{mathpartir}
\usepackage{xspace}
\usepackage{amssymb}
\setcounter{tocdepth}{3}
\usepackage{graphicx}
\usepackage{stmaryrd}
\usepackage{cite}
\usepackage{hyperref}
\usepackage{framed}

@@

\renewcommand\vec[1]{\overline{#1}}

@@

\let\oldneg=\neg
\def\neg{\oldneg\:}

@@

\newcommand{\boolop}{\oplus}
\newcommand{\forallf}[1]{\forall_{\!#1\:}}
\newcommand{\fnull}{\emptyset}
\newcommand{\farg}[1]{a_{#1}}
\newcommand{\vecfarg}[1]{\vec{a}_{#1}}
\newcommand{\fargsort}[1]{\upsilon_{#1}}

@@

%\newcommand{\euf}{\ensuremath{\mathrm{E}}\xspace}
\newcommand{\euf}{\ensuremath{\mathcalx{E}}\xspace}
%\newcommand{\ari}{\ensuremath{\mathrm{A}}\xspace}
\newcommand{\ari}{\ensuremath{\mathcalx{A}}\xspace}

@@

A \emph{signature} $\Sigma$ consists of 
a set $\ssorts{\Sigma}$ of type symbols (or sorts) and
a set $\sfuns{\Sigma}$ of function symbols.
We assume that signatures always include a Boolean type $\Bool$ and constants 
$\ltrue$ (truth) and $\lfalse$ (falsity) of that type,
an infix equality predicate ${\teq} : \tau \times \tau \to \Bool$
for each $\tau \in \ssorts{\Sigma}$,
standard boolean connectives ($\neg$, $\wedge$, $\vee$, etc.),
and a function $\lite : \Bool \times \tau \times \tau \rightarrow \tau$ for each $\tau \in \ssorts{\Sigma}$.
We write $t\typ{\tau}$ to denote a term $t$ of type~$\tau$ and
$\terms( t )$ to denote the set of subterms in $t$.
%We write $\functerms^\Sigma( t )$ to denote the set of $f$-applications in $t$ such that $f \in \sfundefs{\Sigma}$.

A \emph{$\Sigma$-interpretation $\I$} %is a mathematical structure that
maps each $\tau \in \ssorts{\Sigma}$ to a nonempty set $\tau^\I$,
the \emph{domain} of~$\tau$ in $\I$,
and each $\con{f} : \tau_1 \times \cdots \times \tau_n \rightarrow \tau$ in
$\sfuns{\Sigma}$ 
to a total function $\con{f}^\I : \tau_1^\I \times \cdots \times \tau_n^\I \rightarrow \tau^\I$.
A \emph{theory} is a pair $T = (\Sigma, \Mo)$ where 
$\Sigma$ is a signature and  $\Mo$ is a class of $\Sigma$-interpretations,
the \emph{models} of $T$.
A $\Sigma$-formula $\varphi$ is 
\emph{$T$-satisfiable}
if it is satisfied by some interpretation in $\Mo$;
otherwise, it is \emph{$T$-unsatisfiable}.
Formulas $\varphi$ and $\psi$ are \emph{equisatisfiable} if
$\varphi$ is $T$-satisfiable if and only if $\psi$ is $T$-satisfiable.
Given a signature $\Sigma$,
the \emph{theory of equality with uninterpreted functions} \euf
consists of the set of all $\Sigma$-interpretations.
We refer to the type and function symbols in this theory as \emph{uninterpreted}.

%In this paper, we associate quantified formulas with uninterpreted functions (informally, those whose semantics they define).
In this paper, we consider \emph{annotated quantifed formulas} of the form
$\forallf{a} \vec x.\; \varphi$, where the annotation~$a$ is either
an uninterpreted function $\con{f} \in \sfuns{\Sigma}$ or the special symbol
$\fnull$.
Given $\con{f} : \tau_1 \times \ldots \times \tau_n \rightarrow \tau$,
an annotated
quantified formula $\forallf{\con{f}} \vec x.\; \varphi$ is a \emph{function definition}
(\,\emph{for $\con{f}$}\vthinspace) if $\vec x$ is a tuple of variables of type
$\tau_1, \ldots, \tau_n$ and $\forallf{\con{f}} \vec x.\; \varphi$ is equivalent to
$\forall \vec x.\;\, \con{f}( \vec x ) \teq t$ for some $t$. We will write $\forall \vec
x. \varphi$ as shorthand for $\forallf{\fnull} \vec x.\; \varphi$, and $\exists
\vec x. \varphi$ as shorthand for $\neg \forall \vec x.\; \neg \varphi$.

\begin{definition}
A formula $\varphi$ is \emph{well-defined with respect to functions} $\{ f_1, \ldots, f_n \}$ if it is of the form:
\begin{eqnarray} \label{eq:wdf}
(\forallf{f_1} \vec x_1.\; \varphi_1) \wedge \ldots \wedge (\forallf{f_n} \vec x_n.\; \varphi_n) \wedge \psi
\end{eqnarray}
where $f_1, \ldots, f_n$ are not $\fnull$ and distinct, 
and $\psi$ contains no function definitions.
We will commonly call $\psi$ the \emph{conjecture} of $\varphi$.
\ct{Don't we need additional restrictions on the $\varphi_i$'s here?
(E.g., they are quantifier-free, or perhaps existential or containing only let binders.)}
\end{definition}

@@

A set of function definitions $\{ \varphi_1, \ldots, \varphi_n \}$ 
is \emph{well-specified (in $T$)} if $\varphi_1 \wedge \ldots \wedge \varphi_n$ is $T$-satisfiable.
For instance, $\{ \forallf{\con{f}} \vec x.\;\, \con{f}( x ) \teq 0 \}$ is well-specified,
whereas $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) + 1 \}$ is not. 
Note that a well-specified function definition need not specify a terminating function, e.g.
$\{ \forallf{\con{f}} x^\sortint.\;\, \con{f}( x ) \teq \con{f}( x - 1 ) \}$ or even $\{ \forallf{\con{f}} x.\;\, \con{f}( x ) \teq \con{f}( x ) \}$ are well-specified.
\ajr{revise}

@@

Recent techniques~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} 
have focused on finding models for quantified formulas in SMT.
A instantiation-based procedure is shown in~\cite{GeDeM-CAV-09} to be complete for formulas in the \emph{essentially uninterpreted} fragment.
This fragment is limited to universally quantified formulas where all occurrences of variables occur as direct subchildren of uninterpreted functions.
For instance, the formula $\forall x\typ{\sortint}.\;\, \con{f}( x ) \teq g( x ) + 5$ 
is in the essentially uninterpreted fragment since all occurrences of $x$ occur beneath uninterpreted functions $\con{f}$ and $g$.
Other syntactic criteria are identified in this paper that extend this fragment slightly, including cases when variables occur as arguments of arithmetic predicates.
The techniques in~\cite{ReyEtAl-1-RR-13} can be used for finding finite models for quantified formulas over uninterpreted types as well as types having a 
fixed finite interpretation (such as fixed-width bitvectors, datatype enumerations, etc.).
For instance, 
finite model finding techniques can find a model for
the formula $\forall x\typ{\tau} y\typ{\tau}.\;\, \neg x \teq y \vee \con{f}( x ) \teq \con{f}( y )$
where the uninterpreted type $\tau$ is interpreted as a finite set, and $\con{f}$ as injective finite map.

Unfortunately, these fragments do not coincide with the vast majority of quantified formulas that correspond to recursive function definitions.
In particular, the fragments in~\cite{GeDeM-CAV-09} are not applicable when the argument of a recursive function is used in various ways (e.g. within an arithmetic expression), 
while the fragment in~\cite{ReyEtAl-1-RR-13} is not applicable for functions over infinite domains such as the integers.
So, for instance, the quantified formula $\forall x\typ{\sortint}.\;\, \lite( p \leq 0, p( x ) \teq 0, p( x ) \teq 2 * p( x - 1 ) )$ 
is not in the essentially uninterpreted fragment (since $x$ occurs as a child of $-$),
nor does finite model finding apply (since $x$ is of type $\sortint$).

@@

In the remainder of the section, we fix a signature $\Sigma$.
Say that $\Sigma$-formula $\varphi$ is well-defined with respect to functions $\sfundefs{\Sigma}$, 
and the set of function definitions corresponding to $\sfundefs{\Sigma}$ are well-specified.
%We may translate $\varphi$ into a equisatisfiable formula $\psi$ for which known model-finding procedures~\cite{GeDeM-CAV-09, ReyEtAl-1-RR-13} are applicable.
This section presents a method for constructing an extended signature $\extendsig{ \Sigma }$,
and a $\extendsig{ \Sigma }$-formula $\psi$ such that $\varphi$ and $\psi$ are equisatisfiable.
The intuition behind this translation is to use an uninterpreted type \ct{to be defined} to abstract the set of \emph{relevant} tuples for each recursive function $\con{f}$,
and restrict the quantification of the function definition for $\con{f}$ to a variable of this type.
Informally, the relevant tuples $\vec t$ of a function $\con{f}$ are the ones for which the interpretation of $\con{f}( \vec t )$ is relevant to the satisfiability of $\varphi$.

\begin{figure}[t]
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv( \varphi )$ : 
 \begin{itemize}
   \item[] $\mathsf{let}$ $( \psi, D ) = \conv( \varphi, \ptrue )$;
   %\item[] $\mathsf{assert}( D = \emptyset )$;
   \item[] $\mathsf{return}$ $\psi$
  \end{itemize}
\end{enumerate}
\begin{enumerate}
%\begin{framed}
\item[\ ] 
$\conv_0( t\typ{\tau}, p )$ : 
 \begin{itemize}
   \item[] $\mathsf{if}$ $\tau \equiv \Bool$ and $t \equiv t_1 \boolop \ldots \boolop t_n$ :
    \begin{itemize}
      \item[] $\mathsf{let}$ $( r_i, D_i ) = \conv_0( t_i, \pol( \boolop, i, p ) )$ for $i = 1, \ldots, n$;
      \item[] $\mathsf{if}$ $p = \ptrue$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \wedge D_1 \wedge \ldots \wedge D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$ $\mathsf{if}$ $p = \pfalse$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( ( r_1 \boolop \ldots \boolop r_n ) \vee \neg D_1 \vee \ldots \vee \neg D_n, \emptyset )$
      \end{itemize}
      \item[] $\mathsf{else}$
      \begin{itemize}
        \item[] $\mathsf{return}$ $( r_1 \boolop \ldots \boolop r_n, D_1 \cup \ldots \cup D_n )$
      \end{itemize}
    \end{itemize}
  \item[] $\mathsf{else}$ $\mathsf{if}$ $t \equiv \forallf{\con{f}} \vec x.\;\, t_1$
    \begin{itemize}
      \item[] $\mathsf{let} ( r_1, D_1 ) = \conv_0( t_1, p )$;
      \item[] $\mathsf{if}$ $\con{f} \in \sfundefs{\Sigma}$
      \begin{itemize}
        %\item[] $\mathsf{assert}( D_1 = \emptyset )$;
        \item[] $\mathsf{return}$ $( \forall y\typ{\fargsort{\con{f}}}.\;\, ( r_1 [ \vecfarg{\con{f}}( y ) / \vec x ] ), \emptyset )$
      \end{itemize}
      \item[] else
      \begin{itemize}
        \item[] $\mathsf{return}$ $( \forall \vec x.\; r_1, \forall \vec x.\; D_1 )$
      \end{itemize}
    \end{itemize}
   \item[] $\mathsf{else}$
   \begin{itemize}
     \item[] $\mathsf{return}$ $( t, \{ \exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{f}( z ) \teq \vec t ) \mid f( \vec t ) \in \terms( t ), f \in \sfundefs{\Sigma} \} )$
   \end{itemize}
 \end{itemize}
\end{enumerate}
\vspace{-2ex}
\caption{A translation procedure $\conv$ for a $\Sigma$-formula that is well-defined with respect to a set of well-specified function definitions $\sfundefs{\Sigma}$.
In the subprocedure $\conv_0$, the procedure $\pol( \boolop, i, p )$ returns the polarity of the $i^{th}$ child of a $\boolop$-application having polarity $p$,
where $\boolop$ is an interpreted predicate.
}
\label{fig:encoding}
\end{figure}

In more detail, 
say our signature $\Sigma$ contains a set of defined function symbols $\sfundefs{\Sigma} \subseteq \sfuns{\Sigma}$.
We extend the signature of $\Sigma$ to a signature $\extendsig{\Sigma}$ containing the following additional definitions.
For each $\con{f} \in \sfundefs{\Sigma}$ of type $\tau_1 \times \ldots \times \tau_n \rightarrow \tau$, 
the signature $\extendsig{\Sigma}$ includes:
\begin{enumerate}
\item[-] an uninterpreted type $\fargsort{\con{f}}$, and
\item[-] a vector of $n$ uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ of type $(\fargsort{\con{f}} \rightarrow \tau_1)$, $\ldots$, $(\fargsort{\con{f}} \rightarrow \tau_n)$.
\end{enumerate}
%The interpretation of uninterpreted sort $\fargsort{\con{f}}$ will denote the elements (tuples) on which the function $\con{f}$ is applied.
%The role of the uninterpreted functions $\farg{\con{f}}^1$, $\ldots$, $\farg{\con{f}}^n$ will be discussed more in the following.

Given the extended signature, we run the procedure $\conv$ in Figure~\ref{fig:encoding} on $\varphi$.
This calls the subprocedure $\conv_0$, which takes two arguments: the term $t$ to translate, and a \emph{polarity} $p$ (either $\ptrue$, $\pfalse$, or $\pnone$).
It returns a pair of the form $( r, D )$, where $r$ is a term of type $\tau$, and $D$ is a set of formulas.
%The role of $D$ is to ensure that the (restricted) function definition for $\con{f}$ includes certain tuples in its domain, which we explain more in the following.
At a high level, the translation modifies $\varphi$ in two ways.
First, it restricts the quantification on function definitions for $\con{f}$ to the corresponding uninterpreted type $\fargsort{\con{f}}$.
Second, it augments $\varphi$ with additional constraints of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$
which ensure that this restriction for the defintion $\con{f}$ covers all relevant tuples of terms,
namely the ones for which an $\con{f}$-application exists and is relevant to the satisfiability of the current branch of $\varphi$.

In the case that $t$ is an application of (interpreted) predicate $\boolop$ (e.g., a boolean connective, or equality),
we recursively call $\conv$ on its subchildren $t_i$ and polarity $\pol( \boolop, i, p )$, where $\pol$ is defined as\footnote{The negation $\neg p$ of polarity $p$ is 
$\pfalse$ if $p$ is $\ptrue$,
$\ptrue$ if $p$ is $\pfalse$, and
$\pnone$ if $p$ is $\pnone$.}:
\begin{equation*}
\pol( \boolop, i, p ) = \begin{cases}
                         \pnone & \boolop \equiv \teq \text{ or } ( \boolop \equiv \lite \text{ and } i=0 ) \\
                         \neg p & \boolop \equiv \neg \\
                         p & \text{otherwise}
                         \end{cases}
\end{equation*}
If $t$ has positive polarity,
we construct a conjunction of the result of the recursive calls with formulas $D_1, \ldots, D_n$.
Similarly, if $t$ has negative polarity,
we construct a disjunction of the result of the recursive calls with the negation of formulas $D_1, \ldots, D_n$.
It $t$ has no polarity,
then we return the union of $D_1 \cup \ldots \cup D_n$.

In the case that $t$ is a quantified formula $\forallf{\con{f}} \vec x.\; t_1$, it recursively calls $\conv$ on its body with the same polarity.
In the case that $t$ is a function definition of a function in $\sfundefs{\Sigma}$,
we instead construct a quantified formula over a single variable $y$ of type $\fargsort{\con{f}}$,
and replace all occurrences of $\vec x$ in $r_1$ with $\vecfarg{\con{f}}( y )$.
In this case, since function definitions are a top-level conjunct, by case analysis on the return values of $\conv$, we know that $D_1$ is empty.
In the case that $t$ is not a function definition, 
the resulting quantified formula is reconstructed with the term $r_1$ returned by the recursive call,
and a quantified prefix is appended to the formula $D_1$ returned by the recursive call.

Otherwise, $t$ is either an application of an uninterpreted predicate, or a term of a type other than $\Bool$.
In this case, the second component of our return value contains a constraint of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$
for each subterm $\con{f}( \vec t )$ of $t$ such that $\con{f}$ is a function in $\sfundefs{\Sigma}$.
When such constraints are asserted positively, this indicates that the tuple of uninterpreted functions $\vecfarg{\con{f}}$ must include $\vec t$ in its range.

@@

%\begin{lemma}
%Let $\M$ be a model satisfying $\conv( \varphi )$, and
%%and let $\conv( \varphi )^\ast$ be the quantifier-free formula obtained by 
%let $M$ be a set of literals such that $\M \models M$ and $M \models_p \conv( \varphi )$.
%For each assignment $\sigma$ over $\vars{ M }$,
%each $\con{f}( \vec t ) \in \terms( M \sigma )$ where $\con{f} \in \sfundefs{\Sigma}$, 
%there exists a formula of the form $\exists z\typ{\fargsort{\con{f}}}. ( \vecfarg{\con{f}}( z ) \teq \vec t )$ in $M \sigma$.
%\end{lemma}
